\chapter{Goals}

\section{Main goal}

The main objective of the project is to evaluate impact and improve models performance when trained using augmented datasets with synthetic image generation techniques. 

To achieve this goal, the project aimed to develop a pipeline shown in Figure \ref{fig:pipeline_diagram} capable of generating synthetic data using generative models, which could be generalized and enable further enhancements to existing classification models. Through the collaboration with Cogniflow, our research gained valuable resources and guidance. We were granted unlimited access to Cogniflow's platform, simplifying the development and management of our classification models and facilitating the seamless execution, comparison, and tracking of experiments. Additionally, we held two consultative meetings with the Cogniflow team, which provided specific insights for transitioning our development to a production environment should the outcomes be promising. The initial project idea was also conceived jointly with Cogniflow, aligning our research goals from the onset. 

As part of our mutual objective with Cogniflow, we are dedicated to formulating a solution that demonstrates efficiency, scalability, and minimal computational resource demands. Scalability is of particular significance, especially for small businesses, as it empowers them to implement better classification models in a commercial context without the need for substantial infrastructure investments or extensive resource allocation. This scalability should be adaptable to varying data input sizes and computational requirements, not only ensuring versatility but also commercial viability and facilitating the seamless integration of these models, even for enterprises operating within resource constraints.

\subimport{./figures/diagrams}{pipeline.tex}

The introduced pipeline incorporates a parallel path alongside the traditional training steps for a classification model. This includes image sampling, generative model training, image generation, and combination steps, each posing unique challenges and objectives.

\section{Secondary goal: Contributions}

In addition to the mentioned principal objective, we established a distinct goal to enrich public repositories and contribute valuable resources to the broader community. This endeavor not only serves to assist future research in this domain but also extends its potential benefits to other related subjects.
